\section[Concepts of libclang]{Concepts of \libclang}\label{sec:libclangConcepts}

%Translation units,  AST and cursors, traversing the tree, types
Before we discuss the \R{} facilities for working with \libclang, it
is useful to understand the basic concepts of the library.  These
include translation units, abstract syntax trees and cursors, types
and tokens.  \libclang{} exposes other concepts, but these are the
important ones for our purposes.

Source code is arranged in files. A project may be made up of more
than one related files.  For example, we may define related routines
for one task in one file and other routines in a separate file.
Header files are used for declarations and pre-processor definitions
that are shared by several files.  When we parse a file that reads the
source and substitutes the content of any header files referenced by
an \Ckeyword{\#include} directive.  The entity resulting from parsing
the document is called a \textit{translation unit} (TU).

A translation unit contains all the information from the corresponding
source file.  A single TU is often sufficient for our purpose of
exploration and analysis. However, if the code in one translation unit
refers to routines or variables in an other source file, we have to
merge it with another translation unit to resolve those references.
\libclang{} uses an \textit{index} as the container for several
related translation units.  When we want to deal with the translation
units together, we use the same index to parse them and this then
allows \libclang{} and us to connect references to elements across
translation units.

\libclang{} is a parser and represents a translation unit as a parse
tree.  It first breaks the text into tokens during the lexical
analysis step.  This breaks code of the form \Cexpr{dnorm(n, 1, sd)}
into the separate components \dquote{dnorm}, \dquote{(}, \dquote{n},
\dquote{1}, \dquote{sd}, \dquote{)}.  The parsing stage then maps the
tokens into language concepts.  In this example, it identifies a call
expression. Within the call entity is the reference to the routine
being called (\Cfunc{dnorm}) and then the three arguments to the call.
The parenthesis and comma tokens disappear in the parsing stage as we
move from the individual tokens to higher-level semantic meaning of
the tokens.  However, a cursor still has access to its tokens and we
can obtain these for a given cursor. This is important as \libclang{}
doesn't differentiate between different types of binary operators, for
example. Instead, we find the operation (=, +, -, etc.) from the
tokens.

\libclang{} represents the different parse elements via cursor (of
class \Rclass{CXCursor} in \R.  A cursor has a kind that identifies
its nature or what concept it represents. There are many different
kinds for a cursor (enumerated in \Rpkg{RCIndex} via the variable
\Rvar{CXCursorKind}) and they include FunctionDecl, VarDecl, ParmDecl,
CallExpr, IntegerLiteral, ClassDecl MacroDefinition, EnumDecl,
StructDecl.  Hopefully the names indicate the purpose of the cursor.

Along with its kind, a cursor has child cursors.  These are the
components such as the reference to the routine and the parameters in
a call, the fields and methods in a \Cpp{} class or the left and right
hand side of an assignment expression. Each of these child cursors has
a kind and also, possibly, its own sub-cursors. As a result, we have a
tree, or hierarchical structure.  The translation unit is the
container for the top-most or root cursor of the tree.  This root
cursor has the abstract kind \ClangKind{TranslationUnit}.  Its
children are the top-level elements of the source code file,
e.g. global variables, routines, class definitions, pre-processor
terms. Each of these has its own child cursors and so on.

This concept of a cursor tree is important when we examine how to
extract information using the lower-level facilities in
sections~\ref{ref:LowLevel}~and~\ref{sec:Examples}.  We will traverse
the tree, either one node at a time or directly by querying child
cursors and their children.

In addition to cursors, \libclang{} manages type information.  In the
\C{} and \Cpp{} languages, every expression has a type, be it a
variable, call to a routine, binary operator and so on.  \libclang{}
associates a type with each cursor.  Importantly, it ensures that
there is a single type object describing each unique data type in the
translation unit, and across related translation units.  This allows
us to reason about and resolve types easily.

Like a cursor, a type has a kind such as \ClangTypeKind{Int},
\ClangTypeKind{Float}, \ClangTypeKind{Double},
\ClangTypeKind{Typedef}, \ClangTypeKind{Pointer},
\ClangTypeKind{Record} (for a \Ckeyword{struct}),
\ClangTypeKind{Enum}.  A type also has a name, a size and other
characteristics we can query.


\libclang{} provides other features such as diagnostics for parsing
and compiler errors, code completion used in editors, serializing
translation units, precompiled header files, efficient indexing of
translation units.  Some of these are exposed in the \Rpkg{RCIndex}
package. However, none are essential for understanding how to work
with \libclang{} and extracting useful information from source code.

