\subsection{Finding the signature for a foreign function interface call}\label{sec:RffiEG}

In this example, we focus on working with types in \Rpkg{RCIndex}.

As I mentioned in section~\ref{sec:Introduction}, there are two
packages (\Rpkg{Rffi} and \Rpkg{rdyncall}) that allow us to
dynamically invoke arbitrary native routines within an \R{} session
without having to write or compile any additional wrapper code.  Both
require a description of the signature of the routine to be
called. That is, we need the return type and the number and types of
the parameters.  We don't want the user to have to know or specify the
signature, so we want to be able compute it for them.  \Rpkg{rdyncall}
uses GCC-XML to get the information about the routines and generate
the corresponding signatures.  Here we'll show how to do it with
\Rpkg{RCIndex} and directly from the results of \Rfunc{getRoutines}
applied to the source code describing the routine(s) of interest.

Consider the simple  routine
\begin{CCode}
#include <math.h>
int
square_sin(int *val, int len, double *ans)
{
    int i;
    for(i = 0; i < len ; i++) {
	double tmp = sin(val[i]);
	ans[i] = tmp * tmp;
    }
    return(len);
}
\end{CCode}
This takes a vector of integer values and computes the
square of the sin of each element and returns the values
in the array of \Ctype{double} values. It returns the number
of elements it processed.

We'll look at how we can invoke this routine from \R{} using
\Rpkg{rdyncall} and then how we can generate the appropriate signature
programmatically.  Basically, before we can invoke the routine, we
have to compile the \C{} code into a shared
library, load that into \R{} and find the address of the symbol
\Cfunc{square_sin}.  We do this with a shell command
\begin{ShCode}
R CMD SHLIB rffi.c
\end{ShCode}
and the \R{} code
\begin{RCode}
dyn.load("rffi.so")
sym = getNativeSymbolInfo("square_sin")$address
\end{RCode}
%$

Next, we'll create some sample inputs:
\begin{RCode}
x = 1:10
ans = numeric(length(x))
\end{RCode}

To invoke the routine, we use the function \Rfunc{.rdyncall} function
in the \Rpkg{rdyncall} package. 
The call is 
\begin{RCode}
v = .dyncall(sym, "*ii*d)i", x, length(x), ans)
\end{RCode}
\Rfunc{.dyncall} takes care of marshalling the inputs and outputs
between \R{} and the native code.

The strange (and unaesthetic) string \verb+"*ii*d)i"+ is the
description of the signature of the routine.  It indicates that the
first argument (\verb+*i+) is a pointer to one or more \Ctype{int}
values, the second (\verb+i+) is a simple \Ctype{int} value specifying
the number of elements in the first argument, and the third
(\verb+*d+) is a pointer to one or more \Ctype{double} values.  The
\verb+)+ character separates the inputs from the return type and the
latter in this case is \verb+i+ and so an \Ctype{int}.  Of course, we knew this signature
by looking at the code, althought we didn't know the actual form it
took for \Rpkg{rdyncall}.

We'll write a function that uses \libclang{} and \Rpkg{RCIndex} to
create the signature for an arbitrary routine.  We'll assume our
function is passed a \Rclass{FunctionDecl} object returned by
\Rfunc{getRoutines}.  Recall that this contains the parameters and the
return type.  We can easily generate the signature with something
like
\begin{RCode}
makeSig = 
function(r)
{
   args = sapply(r@params, makeType)
   rtype = makeType(r@returnType)
   paste0( paste(args, collapse = ""), rtype)
}
\end{RCode}
The important part of this function is the call to \Rfunc{makeType}
for each type in the signature and that is the function we must implement next.
The \Rfunc{makeType} function is passed  either a \Rclass{CXCursor} or a \Rclass{CXType}.
We want to deal with the \Rclass{CXType} so must convert a
\Rclass{CXCursor} first. (We could of course use S3 or S4 methods to
do this also.)
\begin{RCode}
   if(is(ty, "CXCursor"))
     ty = getType(ty)
\end{RCode}
Once we have the \Rclass{CXType}, we can query the kind of type it is
with \Rfunc{getTypeKind}.
Given the kind, we first check to see if it corresponds to one of the
primitive \C{} types. We make a list of these by combining
the different enumeration values for \Rclass{CXType}, e.g.,
\Rvar{CXType_Double}, \Rvar{CXType_Int} and so on.
We can do this with
\begin{RCode}
BasicKinds = CXTypeKind[3:24]
\end{RCode}
since the primitive kinds are ordered contiguously in the
enumeration.

If the type does correspond to a basic type, we map it to 
the string that \Rfunc{.rdyncall} expects for this type.
These are listed in a table in the help page for \Rfunc{.rdyncall}.
\dquote{i} corresonds to \Rvar{CXType_Int}, \dquote{I} corresponds
to \Rvar{CXType_UInt} for an unsigned integer, and so on.
We express this map via the named \R{} vector
\begin{RCode}
rdyncallMap = c(
  v = CXType_Void,
  B = CXType_Bool,
  C = CXType_Char_U,
  C = CXType_UChar,
  c = CXType_Char32,
  c = CXType_Char_S,
  S = CXType_UShort,
  I = CXType_UInt,
  J = CXType_ULong,
  L = CXType_ULongLong,
  s = CXType_Short,
  i = CXType_Int,
  l = CXType_Long,
  l = CXType_LongLong,
  f = CXType_Float,
  d = CXType_Double
)
\end{RCode}
The value in the vector is the \libclang{} type kind. The
corresponding name is the \Rpkg{rdyncall} type identifier.
We can match the \libclang{} kind to the values in the vector and then
return the name of the corresponding element.
We do this with the simple function
\begin{RCode}
mapBasicKindToRdyncall =
function(kind)
{
  i = match(kind, rdyncallMap)
  if(is.na(i))
    stop("don't recognize that type")

  names(rdyncallMap)[i]
}
\end{RCode}
  
If the type kind of the type passed to \Rfunc{makeType} is not a basic
type, we have to examine it further.  We can compare the kind to
\Rvar{CXType_Pointer} to determine if it is a pointer, or call
\Rfunc{isPointerType} with the actual type object.  If the type is a
pointer type, we can get the type of element it points to with
\Rfunc{getPointeeType}.  Here we have to treat the \C{} string type
\Cexpr{char *} specially since \Rpkg{rdyncall} does.  We check if the
pointee type is of kind \Rvar{CXType_Char_S} and if it is, we return
the \Rpkg{rdyncall} string \dquote{Z}.

If the pointee type is not a \Rvar{CXType_Char_S}, then we check to
see if it is a basic type again. If so, we will combine the type
string for this basic type with a $\ast$ prefix, e.g., \dquote{*i}.  If
the type is not a basic type, then it is a pointer to a generic \C{}
type and we return \dquote{p}, as \Rpkg{rdyncall} requires.

We also have to deal with types that are not simply basic or pointer
types. For example, we may have a type declaration such as
\begin{CCode}
typedef long VecSize;
\end{CCode}
This is simply an alias or a new name for an existing type.
\Rpkg{rdyncall} doesn't care about the name, but only the actual type.
To deal with this in \Rpkg{RCIndex}, we can
use the canonical type to get at the actual type of the alias 
and pass this to a call to \Rfunc{makeType}.

The entire \Rfunc{makeType} function is defined as
\begin{RCode}
makeType = 
function(ty)
{
   if(is(ty, "CXCursor"))
     ty = getType(ty)

   kind = getTypeKind(ty) # or ty$kind
   
   if(kind %in% BasicKinds) {
     
      mapBasicKindToRdyncall(kind)

   } else if(kind == CXType_Pointer) {
     
     pty = getPointeeType(ty)
     if(getTypeKind(pty) == c(CXType_Char_S))
        "Z"
     else if((pk <- getTypeKind(pty)) %in% BasicKinds)
        sprintf("*%s", mapBasicKindToRdyncall(pk))
     else
        "p"
    
   } else if(kind == CXType_Typedef) {
     makeType(getCanonicalType(ty))
   }
}
\end{RCode}
%$

With the function defined, we can use it on
the routines in  \file{rffi.c}:
\begin{RCode}
r = getRoutines(system.file("exampleCode", "rffi.c", package = "RCIndex"))
sapply(r, makeSig)
\end{RCode}
\begin{ROutput}
square_sin    string1    string2   typedefn 
 "*ii*d)i"      "Z)v"      "Z)v"      "l)v" 
\end{ROutput}
These correspond to the declarations
\begin{CCode}
int square_sin(int *val, int len, double *ans);
void string1(char *x);
void string2(const char *x);
void typedefn(VecSize s)
\end{CCode}
The final signature illustrates mapping our
\Ckeyword{typedef} example to the  \Rpkg{rdyncall} string \dquote{l}
corresponding to a \Ctype{long}.

%We can identify whether a parameter is identified as being immutable.
%This allows us to avoid copying the argument before we pass it to
%the native routine.

We could go further with generating descriptions for use with
\Rpkg{rdyncall} or \Rpkg{Rffi} to process \Ckeyword{struct} and
\Ckeyword{union} types and \Cpp{} classes and so on.

